FastSRM: A fast, memory efficient and identifiable implementation of the shared response model

### 2022-08-03 paper/main.tex
update intro

\label{prop:optimaldetsrm}
\end{proposition}
The proof is available in appendix~\ref{proof:deterministic}.
In the case of probabilistic SRM we can obtain very similar results.
%
However the
algorithm applied on reduced data needs to be slightly modified.
%
We have the following result:
\begin{proposition}[Optimal spatial decomposition for probabilistic FastSRM]
  Let $(A_i)_i, \sigma_i, \Sigma_s$ be the solution obtained by probabilistic SRM on data
  $\xb_i \in \RR^v$ and $(A_i')_i, \sigma_i', \Sigma_s'$ the solution obtained by ProbSRM on
  data $\zb_i = U_{\xb_i}^{\top} \xb_i$, $\zb_i \in \RR^n$ but where updates:
  \begin{align}
    \sigma_i^2 \leftarrow \frac1{n} (\EE[\| \zb_i - A_i \EE[\sbb|\zb]\|^2] + \| \diag(\VV[\sbb | \zb]) \|^2)
  \end{align}
  are replaced by updates
  \begin{align}
    \sigma_i^2 \leftarrow \frac1{v} (\EE[\| \zb_i - A_i \EE[\sbb|\zb]\|^2] + \| \diag(\VV[\sbb | \zb]) \|^2)
  \end{align}
  %
  where $t$ is the number of samples and $v$ is the number of voxels.

@@

We consider SRM as a well principled formulation of the hyperalignment
method~\cite{haxby2011common}.
%
Many methods exist to perform hyperalignment: deep
hyperalignment~\cite{yousefnezhad2017deep} , robust
SRM~\cite{turek2018capturing} , multi-view canonical correlation
analysis~\cite{li2009joint}, ShICA~\cite{richard2021model},
MultiViewICA~\cite{richard2020modeling} optimal
transport~\cite{bazeille2019local} and many more.
%
SRM has been introduced after hyper-alignment in~\cite{chen2015reduced} and has been shown to outperform hyperalignment in many different settings. A recent work has come to a similar conclusion~\cite{bazeille2021empirical}. In this article, we do not discuss further the merit of FastSRM with respect to other methods. We simply observe that SRM is widely used and aim to provide
a faster, more memory efficient and identifiable implementation.  In
our view, FastSRM should be used as a plugin replacement for SRM.

@@

We empirically see that the optimal approach is equivalent to using no spatial decompositions in
terms of performance.
%
This is predicted by the theory in section~\ref{sec:optimal_atlas},
where we demonstrate that these two algorithms yield exactly the same
output from a given input and initialization.
%
% %
% Among approximate methods, the projection
% method yields the best results but are quite far away from optimal results.
%
% %
In general probabilistic methods give much better results than their deterministic
counterpart.
%
This shows the superiority of likelihood based methods.
%
In terms of fitting time, FastSRM is about a thousand time faster than SRM after
100 iterations.
%
When no spatial decomposition is used, the number
of iterations has a very strong impact on performance while it has a small impact
when an optimal spatial decomposition is used.
%
Lastly, looking at the convergence curves, we see that even after $100$ iterations, algorithms did
not fully converge.
%
This means that in practice a larger number of
iterations is needed, which would yield an even higher difference in fitting time
between methods using no spatial decomposition and methods using an optimal
spatial decomposition.
%

@@

\subsection{Identifiability increases stability}
Assuming that the data follow the model, identifiability ensures that the
problem is well-posed since the solutions are well-characterized. Note however that SRM
remains non-convex so there is no guarantee that the optimal solution is found.
%
% \bt{I find this a bit ambiguous: you give te impression that you gurantee that the optimal solution is used. Rephrase.}
%
In real fMRI data, the model cannot be expected to hold perfectly, but we can
hope for greater stability in the parameters recovered than if an non-identifiable
model is used.

To measure the stability of the shared components obtained from the Sherlock
dataset, we divide the subjects into two roughly equal groups and extract the
common components in each group.
%
The components are then matched using the Hungarian algorithm and the stability
measure is obtained from the average correlation of the matched components.
%
The procedure is repeated 9 times using the Brainiak implementation (that is not
identifiable since the shared components covariance is unconstrained) and our
FastSRM implementation.
%
We plot the results as a scatter plot in
Figure~\ref{exp:identifiability}, where, for each repetition, the
x-axis indicates the stability measure obtained with Brainiak's
implementation and the y-axis the stability measure obtained with
FastSRM (identifiable since the shared components covariance is
diagonal).
%
We see that introducing a diagonal source covariance improves stability.
%

@@

FastSRM inherits from all the weaknesses of SRM including the fact that
mixing matrices are assumed to be orthogonal, which is debattable in practice.
Another desirable property of such decompositions is the spatial smoothness of the components. While it is often observed that SRM components are smooth, nothing imposes such a constraint in the model. We highlight that the optimal spatial decomposition is only optimal for the model we consider. Whether this representation is interpretable or not is irrelevant as it is only an intermediate step produced used during training.

%
It is an open question whether dimension reduction and learning of the shared
components could be done jointly and efficiently without assuming orthogonal
mixing matrices.
%
Our approach based on optimal spatial decomposition shows that SRM-like methods do not need the
full data to provide accurate estimates of shared components. We believe that such
insights may guide the design of future methods.

>>>>

\label{prop:optimaldetsrm}
\end{proposition}
The proof is available in appendix~\ref{proof:deterministic}.
In the case of probabilistic SRM we can obtain very similar results.
%
However the
algorithm applied on reduced data needs to be slightly modified.
%
We have the following result:
\begin{proposition}[Optimal spatial decomposition for probabilistic FastSRM]
  Let $(A_i)_i, \sigma_i, \Sigma_s$ be the solution obtained by probabilistic SRM on data
  $\xb_i \in \RR^v$ and $(A_i')_i, \sigma_i', \Sigma_s'$ the solution obtained by ProbSRM on
  data $\zb_i = U_{\xb_i}^{\top} \xb_i$, $\zb_i \in \RR^n$ but where updates:
  \begin{align}
    \sigma_i^2 \leftarrow \frac1{n} (\EE[\| \zb_i - A_i \EE[\sbb|\zb]\|^2] + \| \diag(\VV[\sbb | \zb]) \|^2)
  \end{align}
  are replaced by updates
  \begin{align}
    \sigma_i^2 \leftarrow \frac1{v} (\EE[\| \zb_i - A_i \EE[\sbb|\zb]\|^2] + \| \diag(\VV[\sbb | \zb]) \|^2)
  \end{align}
  %
  where $n$ is the number of samples and $v$ is the number of voxels.

@@

We consider SRM as a well principled formulation of the hyperalignment
method~\cite{haxby2011common}.
%
Many methods exist to perform hyperalignment: deep
hyperalignment~\cite{yousefnezhad2017deep} , robust
SRM~\cite{turek2018capturing} , multi-view canonical correlation
analysis~\cite{li2009joint}, ShICA~\cite{richard2021model},
MultiViewICA~\cite{richard2020modeling} optimal
transport~\cite{bazeille2019local} and many more.
%
SRM has been introduced after hyper-alignment in~\cite{chen2015reduced} and has been shown to outperform hyperalignment in many different settings. Specifically, SRM is natively a group model, which hyperalignment is not, it is more computationally efficient, with the avoidance of searchlight loops, and sometimes obtains higher accuracy than Procrustes hyperalignment~\cite{bazeille2021empirical}. In this article, we do not discuss further the merit of FastSRM with respect to other methods. We simply observe that SRM is widely used and aim to provide a faster, more memory efficient and identifiable implementation.  In our view, FastSRM should be used as a plugin replacement for SRM.

@@

We empirically see that the optimal approach is equivalent to using no spatial decompositions in
terms of performance.
%
This is predicted by the theory in section~\ref{sec:optimal_atlas},
where we demonstrate that these two algorithms yield exactly the same
output from a given input and initialization.
%
% %
% Among approximate methods, the projection
% method yields the best results but are quite far away from optimal results.
%
% %
We also see that ProbSRM gives much better results than DetSRM.
%
In terms of fitting time, FastSRM is about a thousand time faster than SRM after
100 iterations.
%
When no spatial decomposition is used, the number
of iterations has a very strong impact on performance while it has a small impact
when an optimal spatial decomposition is used.
%
Lastly, looking at the convergence curves, we see that even after $100$ iterations, algorithms did
not fully converge.
%
This means that in practice a larger number of
iterations is needed, which would yield an even higher difference in fitting time
between methods using no spatial decomposition and methods using an optimal
spatial decomposition.
%

@@

\subsection{Identifiability increases stability}
Assuming that the data follow the model, identifiability ensures that the
problem is well-posed since the solutions are well-characterized. More precisely, if the data are generated according to an identifiable model, in the limit of infinite samples, a perfect optimization algorithm is guaranteed to recover the parameters used to generate the data. In contrast, in an unidentifiable model, there are more than one set of parameters that can generate the same data making the search for the good parameters ill-defined. In practice, unidentifiable models have a more complex optimization landscape with many more local minima than identifiable models (any local minima in the identifiable model produces as many local minimas as they are indeterminacies in the corresponding unidentifiable model). This is why the unidentifiable models are therefore expected to be more sensitive to initialization and less robust to small changes in the data than their identifiable counterparts. Note however that SRM
remains non-convex so there is no guarantee that the optimal solution is found.

%
% \bt{I find this a bit ambiguous: you give te impression that you gurantee that the optimal solution is used. Rephrase.}
%
In real fMRI data, the model cannot be expected to hold perfectly, but we can
hope for greater stability in the parameters recovered than if an non-identifiable
model is used.

To measure the stability of the shared components obtained from the Sherlock
dataset, we divide the subjects into two roughly equal groups and extract the
common components in each group.
%
The components are then matched using the Hungarian algorithm~\cite{kuhn1955hungarian} and the stability
measure is obtained from the average correlation of the matched components.
%
The procedure is repeated 9 times using the Brainiak implementation (that is not
identifiable since the shared components covariance is unconstrained) and our
FastSRM implementation.
%
We plot the results as a scatter plot in
Figure~\ref{exp:identifiability}, where, for each repetition, the
x-axis indicates the stability measure obtained with Brainiak's
implementation and the y-axis the stability measure obtained with
FastSRM (identifiable since the shared components covariance is
diagonal).
%
We see that introducing a diagonal source covariance improves stability.
%

@@

FastSRM inherits from all the weaknesses of SRM including the fact that
mixing matrices are assumed to be orthogonal, which is debattable in practice.
Another desirable property of such decompositions is the spatial smoothness of the components. While it is often observed that SRM components are smooth, nothing imposes such a constraint in the model. We highlight that the optimal spatial decomposition is only optimal with respect to the SRM model. More precisely, it gives a reduced representation such that applying the SRM model on the reduced representation or on the full data gives the same result. Whether this representation is interpretable or not is irrelevant to us, since it is only a means to accelerate the SRM algorithm. However, as this representation is not unique, it may be interesting to find the “most interpretable” optimal spatial decomposition. Note that this would not change the end result of the SRM algorithm but could be of interest for interpretation purposes.
%
It is an open question whether dimension reduction and learning of the shared
components could be done jointly and efficiently without assuming orthogonal
mixing matrices.
%
Our approach based on optimal spatial decomposition shows that SRM-like methods do not need the
full data to provide accurate estimates of shared components. We believe that such
insights may guide the design of future methods.
